Enabling high resolution hydrologic routing with machine learning assisted waterbody classification

### 2024-03-25 manuscript/manuscript.tex
word/character limits

\vspace{-2em}
\section*{Keypoints}
\begin{enumerate}
	\item We demonstrate an automated deep learning pipeline for waterbody database creation that includes a novel waterbody classifier with an accuracy rate approaching 80\%
	\item The pipeline produced similar results in terms of raw waterbody counts, area distributions, and waterbody types as that of a much more expensive to produce hand-curated product
	\item The modular design of the pipeline makes it future-proof to forthcoming data products and algorithms as well as easily tailored to a variety of hydrologic routing tasks	
\end{enumerate}
\newpage

\section*{Abstract}	
	Some waterbody databases are focused on delineating all areas of standing water irrespective of the persistence or “type” of the waterbody. However, for other applications, such as integration of waterbodies into hydrologic routing schemes, capturing the long-term dynamics of persistent waterbodies is of critical importance. The creation of such databases is costly, often requiring manual intervention in many of the underlying steps. As an alternative, we developed a machine learning pipeline that facilitates the rapid, accurate, and inexpensive creation of these databases using deep neural networks operating on high resolution imagery. In addition to basic delineation tasks, capturing time dynamics, and detection of waterbody boundaries, the pipeline includes a novel approach to classify and prune spurious and ephemeral non-lake waterbodies. We demonstrate the accuracy, performance, scaling, and consistency of the pipeline compared with a much more expensive to produce hand-curated product with the aim of scaling to continental and global extents.

>>>>

\vspace{-2em}
\section*{Keypoints}
\begin{enumerate}
	\item We demonstrate a deep learning pipeline for waterbody database creation with a novel waterbody classifier approaching 80\% accuracy
	\item It produced similar results in terms of waterbody count, surface area, and types as a much more expensive to produce hand-curated product
	\item The modular design of the pipeline is future-proof to forthcoming data products and is easily tailored to specific hydrologic routing tasks
\end{enumerate}
\newpage

\section*{Abstract}	
	Some waterbody databases are focused on delineating all areas of standing water irrespective of the persistence or “type” of the waterbody. However, for other applications, such as integration of waterbodies into hydrologic routing schemes, capturing the long-term dynamics of persistent waterbodies is of critical importance. The creation of such databases is costly, often requiring manual intervention in many of the underlying steps. As an alternative, we developed a machine learning pipeline that facilitates the rapid, accurate, and inexpensive creation of these databases using deep neural networks operating on high resolution imagery. In addition to basic delineation tasks, capturing time dynamics, and detection of waterbody boundaries, the pipeline includes a novel approach to classify and prune spurious and ephemeral non-lake waterbodies. We demonstrate the accuracy, performance, and consistency of the pipeline compared with a much more expensive to produce hand-curated product with the aim of scaling to continental and global extents.
